Nonlinear Intrinsic Variables and State Reconstruction in Multiscale Simulations

### 2013-12-27 paper/revision2.tex
Added files related to paper

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
%
These developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by an affine transformation
(see, by analogy, the discussion in Lafon {\em et al.}\cite{lafon2006data}),
the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information).
%
Clearly, the ability to systematically construct a {\em unique} and {\em consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities, is invaluable.
%
We will call these coordinates Nonlinear Intrinsic Variables.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements in a test data set from a training data set
that consists of {\em different} observations.
%
To construct this empirical mapping and the associated observers,
accurate interpolation tools must be available in the embedding space; to this
end, we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

@@

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
We impose two critical assumptions. First, the measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t)),
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold,
and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables (with $d \ll n$).
%
Second, the dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
	\label{eq:dynamics}
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
%
The independence is our second critical assumption.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$ \cite{singer2008non}.
%
Because the empirical method we will describe is independent of the observation function $\mathbf{f}$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$
(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
%
We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model.

@@

\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
\begin{equation}
	d x_i(t) = - \nabla U_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where the drift term is expressed as a gradient of a potential field $U_i$. Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
\begin{equation}
	\mathcal{L} g = \Delta g - \nabla U \cdot \nabla g
	\label{eq:limiting_operator}
\end{equation}
where $g:\mathcal{M}_x \rightarrow \mathbb{R}$ is any real function on the domain of the underlying variables.
Since we assume independent underlying variables, the potential is additive \cite{singer2008non}
\begin{equation}
	U(\mathbf{x}(t)) = \sum \limits _{i=1}^d U_i(x_i(t)).
\end{equation}

This converges implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator.
%
As such, according to \eqref{eq:limiting_operator}, they depend on the potential field.
%
Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are sampled from the same common region.
%
For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation samples the same underling free energy surface.

In order to merge observations from different regions, we need to account for the different potential fields.
%
 This can be achieved by a different normalization of the kernel. In case the normalized kernel is defined as $\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1}$, the limiting diffusion operator is the Laplace-Beltram operator \cite{nadler2006diffusion}, given by
\begin{equation}
	\widetilde{\mathcal{L}} g = \Delta g.
\end{equation}
%
In other words, this diffusion captures only the geometry of the observations and invariant to the different potential fields.
%
Therefore, this normalization separates the geometry of the underlying manifold of intrinsic variables from the statistics of the observations.

Unfortunately, in practice, this normalization is not sufficient, and one needs to consider various affine transformations to relate different sets of intrinsic variables.
%
This has been explored in Lafon {\em et al.} \cite{lafon2006data}, but is currently beyond the scope of this paper.
%
We will only consider applications where diferent simulations sample the same underlying potential field.

@@

@@

For a brief comparison of LP over other reconstruction techniques, we also reconstruct configurations from the
NIV components using simple nearest-neighbor interpolation.
%
The average reconstruction error, scaled by the average bond length within the molecule, is shown in Figure \ref{fig:ala_mse};
LP arguably outperforms simple nearest neighbor search for all of the atoms.

@@

>>>>

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
%
Often, high-dimensional data do not uniformly fill the entire ambient space, but rather lie on a lower-dimensional manifold.
%
Discovering a parameterization of the manifold therefore yields a compact representation of the data, and may lead to efficient analysis and processing schemes.
%
In particular, these developments have the potential to significantly enable the computational exploration
of physicochemical problems.
%
Traditionally, state reduction in chemical systems has been done using some {\em a priori} knowledge about separation of timescales,
for example, using the quasi-steady state approximation to reduce a network of chemical reactions \cite{....}.
%
Alternatively, empirical knowledge about the system of interest has been used to develop coarse grained descriptions, 
such as grouping atoms of a macromolecule into larger ``beads'' to accelerate simulations \cite{...}.
%
The use of manifold learning allows for a systematic discovery of a data-driven, low-dimensional description that circumvents the need for such {\em a priori} knowledge of the system.
%
If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.
%
For example, such reduction coordinates may be the dihedral angles inferred from trajectories of the atoms of a macromolecule.

While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
 data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
different data sets on the same low-dimensional hyperplane in the ambient space
will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
%
While this can be easily rectified by an affine transformation
(see, by analogy, the discussion in Lafon {\em et al.}\cite{lafon2006data}),
the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data sets are obtained using different instrumental modalities
(such as when one wants to merge molecular dynamics data with, for example, spectral information).
%

In this paper, we propose a new method to embed data in a low-dimensional space.
%
This method is based on a kernel constructed from a Riemannian metric that is specially designed to exploit the variability of the data in local neighborhoods.
%
By solving the appropriate eigenvector problem, local affinities from the kernel are integrated into a global coordinate system for the data.

We will demonstrate that this produces a {\em unique} and {\em consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities.
%
We will call these coordinates Nonlinear Intrinsic Variables.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements in a test data set from a training data set
that consists of {\em different} observations.
%
To construct this empirical mapping and the associated observers,
accurate interpolation tools must be available in the embedding space; to this
end, we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.

@@

\subsection{Problem Formulation}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
%
We impose two critical assumptions. First, the measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t)),
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold,
and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables (with $d \ll n$).
%
Second, the dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
	\label{eq:dynamics}
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
%
The independence is our second critical assumption.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$ \cite{singer2008non}.
%
Because the empirical method we will describe is independent of the observation function $\mathbf{f}$,
we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
%
The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$
(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
%
We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model;
this is further discussed and demonstrated in the experimental study.

@@

\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
\begin{equation} \label{eq:Langevin}
	d x_i(t) = - \nabla U_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where the drift term is expressed as a gradient of a potential field $U_i$. Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
\begin{equation}
	\mathcal{L} g = \Delta g - \nabla U \cdot \nabla g
	\label{eq:limiting_operator}
\end{equation}
where $g:\mathcal{M}_x \rightarrow \mathbb{R}$ is any real function on the domain of the underlying variables.
Since we assume independent underlying variables, the potential is additive \cite{singer2008non}
\begin{equation}
	U(\mathbf{x}(t)) = \sum \limits _{i=1}^d U_i(x_i(t)).
\end{equation}
We remark that, unlike in diffusion maps \cite{...}, when the limiting operator is in the domain of the observations $\mathbf{Y}$, here the Laplacian operator that is based on the Mahalanobis distance converges to a limiting operator in the domain of the intrinsic variables $\mathbf{x}$.

The convergence to the limiting operator implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator.
%
As such, according to \eqref{eq:limiting_operator}, they depend on the potential field.
%
Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are sampled from the same common region.
%
For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation sufficiently samples the same underling free energy surface.

In order to merge observations from different regions, we need to account for the potential field.
%
The fundamental question is, to what extent the density of sampling different regions affects the coordinates we obtain.
%
In principal, we can decouple diffusion dynamics from geometry by normalizing the diffusion kernel with sample density.
%
This has been explored for diffusion maps independently of NIV.
%
When the normalized kernel is defined as $\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1}$, the limiting diffusion operator is the Laplace-Beltrami operator \cite{nadler2006diffusion}, given by
\begin{equation}
	\widetilde{\mathcal{L}} g = \Delta g.
\end{equation}
%
In other words, this diffusion captures only the geometry of the observations and is invariant to the potential field.
%
Therefore, this normalization separates the geometry of the underlying manifold of intrinsic variables from the statistics of the observations.

Since local coordinates are based on the noise term in \eqref{eq:Langevin}, sets of measurements from different regions
will essentially detect the same system of coordinates (the same variable ``axes"), yet the different sampling densities arising from the drift term in \eqref{eq:Langevin} will
result in different scalings of these axes.
%
Applying the Laplace-Beltrami normalization in the NIV context will eliminate the sampling density effects.
%
An alternative approach would be to estimate directly the drift term for each intrinsic variable by finding the optimal transformation that would
align the common regions (in the spirit of \cite{...}).

@@

To determine whether a sufficient amount of data is available to obtain an accurate embedding, we compute the NIV embeddings for different numbers of data points 
(arising from different numbers of trajectories initialized randomly in the full concentration space).
%
Figure \ref{fig:rxn_convergence} shows the two-dimensional NIV embeddings and corresponding eigenvalue spectra of the kernel for the Gillespie SSA example computed from 10, 100, 500, and 1500 data points.
%
We observe that the eigenvalue spectra appear converged for 100 data points. 
%
This implies that a sufficient amount of data is available merely to discover the intrinsic variables.
%
However, we observe that 100 points are insufficient to construct a consistent embedding.
%
In order to account for the different scaling of the NIV coordinate axes, we require at least 500 data points.
%
This last convergence is exemplified in Figure \ref{fig:rxn_convergence} (c) and (d).
%
The convergence of the spectrum of the kernel and the consistency of the NIV embedding is an empirical indicator that a sufficient amount of data is available.
\begin{figure*}[ht]
    \includegraphics[width=6in]{make_figures4b.ps}
    \caption{Two-dimensional NIV embeddings (top) and corresponding eigenvalue spectra (bottom) calculated from subsets of $\mathcal{Y}_1$ containing (a) 10 data points, (b) 100 data points, (c) 500 data points, and (d) 1500 data points. One can see that the embedding and spectrum in (a) is inconsistent with the other embeddings. }
    \label{fig:rxn_convergence}
\end{figure*}

@@

For a brief comparison of LP over other reconstruction techniques, we also reconstruct configurations from the
NIV components using simple nearest-neighbor interpolation.
%
The average reconstruction error, scaled by the average bond length within the molecule, is shown in Figure \ref{fig:ala_mse};
LP arguably outperforms simple nearest neighbor search for all of the atoms.
%
To determine why LP outperforms nearest neighbor search, we look at the errors at different scales in the LP algorithm.
%
Figure \ref{fig:LapPyr_ala2_errors} shows the true and reconstructed positions, as well as the error, for the z-coordinate of atom 10 as a function of $\psi_1$ at Scale 8 in the LP algorithm.
%
One can see that the error is larger in some regions (when $\psi_1$ is large) and smaller in other regions (when $\psi_1$

@@

\begin{figure*}[ht]
    \centering
    \includegraphics[width=3in]{fig10b}
    \caption{(a) True position (in blue) and reconstructed position (in black) of the z-coordinate of atom 10 as a function of $\psi_1$. (b) The error in the reconstructed position of the z-coordinate of atom 10 as a function of $\psi_1$. One can see that the error is larger when $\psi_1$ is large, demonstrating that our data is indeed multiscale and that Laplacian Pyramids is an appropriate interpolation algorithm to use.}
    \label{fig:LapPyr_ala2_errors}
\end{figure*}
